### 2014-07-09 tex/main-body-text.tex
Edits during final read through.

\subsection*{{\sc flock} as a special case of {\sc structure}}
The goal of {\sc structure}'s MCMC algorithm is to sample from the space of 
$Z$'s in proportion to their posterior probability.  If, all that was desired
was a point estimate of the $Z_i$'s that maximized the posterior probability,
then one standard approach to finding that maximimum {\em a posteriori}, or MAP, 
estimate, would be to use simulated annealing \citep{Kirkpatricketal1983}.
To do simulated annealing in the {\sc structure} model with no admixture and
no correlated allele frequency prior, after integrating out the unknown
allele frequencies, the updates for each $Z_i$ would be made from the 
distribution
\[
\biggl[P(Z_i=k~|~Y_i, ~Z_{(-i)},~Y_{(-i)\ell})\biggr]^\beta
\]
which is the same full conditional distribution as the marginalized {\sc structure}
model (equation~\ref{eq:fc}), raised
to the power $\beta$.  In typical applications of simulated annealing,
$\beta$ is set to a starting value less than 1, and, as the algorithm proceeds,
$\beta$ is increased according to a ``cooling schedule''\citep{Hajek1988}, until
the algorithm finds a local maximum in the posterior probability surface.  
When $\beta$ is small, the process can easily traverse ``valleys''
in the posterior probability surface and has a better chance of searching
more of the space for the maximum value.  As $\beta$ gets larger, it becomes
more probable that the process will
move uphill on the probability surface, tending toward the local
maximum near where it currently is.

@@

For each set of runs, {\sc flock} reports the 
normalized likelihood values (the exponentiated log-likelihood value for an individual's 
membership to each of $K$ clusters, scaled to sum to one) for a run belonging to the longest plateau - refered to as the 'best run' by \citep{Duc&Tur2012} (see below).
This value is comparable to the $q_i$ values
returned by {\sc structure}, which, in the no-admixture case, are posterior probabilities of cluster membership.
The program {\sc clumpp} \citep{Jak&Ros2007} was used to relabel clusters among different runs,
while the program {\sc district} \citep{Rosenberg2004} was used to plot individual \textit{$q_i$} values.

\citet{Duc&Tur2012} promote a method for using {\sc flock} to estimate, $K$, the number of
subpopulations. The user must specify both the number of runs (different initial allocations of individuals 
to each cluster) as well as the number of iterations. (In every iteration each individual has the opportunity to be reallocated 
to a different cluster.) At the conclusion of each run, the log-likelihood value for membership to all $K$ clusters is computed
and the log likelihood difference (LLOD), the difference between the cluster with the highest likelihood 
and second highest likelihood, is calculated for each individual. {\sc flock} identifies runs that 
have converged to the same solution by detecting that they have the same mean LLOD score, and, in 
{\sc flock} parlance, a series of runs that have converged to the same solution are termed a ``plateau.'' 
The plateau record is a list of how many runs converged to the same allocation
of individuals. 
\citet{Duc&Tur2012} developed two stopping rules that they use to estimate $K$. In the absence of 
strong prior beliefs about the value of $K$, \citet{Duc&Tur2012} advocate running  {\sc flock} until either a single plateau 
$\geq 6$  is reached or four consecutive $K$ values yield no plateaus. If the first stopping condition is satisfied,
the $K$ for which there is a single plateau $\geq 6$ is the point estimate for $K$. Alternatively, 
if multiple plateaus are observed with at least one plateau $\geq 6$, the largest value of $K$ that satisfies 
this conditon should be taken as the lower bound estimate of $K$. If no plateaus $\geq 6$ are observed, $K$
is \textit{undecided}.

@@

\subsection*{Results} 
Run times for both programs were relatively short, albeit non-trivial. All runs of {\sc flock} 
completed in 1133 minutes. Each of the six groups of runs averaged 188.8 minutes.
All runs of {\sc structure} were completed in 
486 minutes and averaged 80.8 minutes to complete each of the six groups of runs.
The processing time of each algorithm increased as the number or 
putative clusters increases. Increasing
$K$ from of 2 to 6 increased the run time for {\sc flock} by a factor of 2.4, while the run time 
for {\sc structure} increased by a factor of 1.6.

{\sc flock} and {\sc structure} provide similar results in 
terms of how individuals are allocated among runs at low values of $K$, 
while at larger values, {\sc flock} often found, in addition to the solution identified by {\sc structure}, 
multiple other solutions. For all six groups 
of runs for $K = 2$ to 4, {\sc flock} resulted in nearly identical clustering of individuals 
(See supplementary materials) none of which differed substantially from results obtain from 
{\sc structure} (Fig 1). Of the 'best runs' for the six groups for $K$ = 2, two  
runs yielded identical clustering (same mean LLOD score) while all other runs 
differed by at most the allocation of 13 individuals. As 
$K$ increases so do the number of solutions
that {\sc flock} found (Fig 1). For all groups of runs for $K$ $\geq 2$, no plateaus were observed - all intial 
allocations resulted in solutions that differed. 
{\sc flock} still produces estimates of the LLOD by individual for the 
'best run', however, it is unclear in these cases what constitutes the 'best run' as all plateau lengths are 
equal to one. Many of these solutions are, however, suprisingly similar to
 the partions identified by {\sc structure}. Both programs idenfity clusters 
which can be attributed to the geographical boundaries identified by \citet{Garzaetal_norcal},
although less reliably so in the case of {\sc flock}.

@@

To investigate the ability of the arrangement of individuals within the reference 
populations to move around the partition space we varied the number of iterations that 
each run goes through to find the 'optimal' allocation of individuals. Unfortuantely 
only the reallocation matrix for the 'best run' is given as output. Increasing the 
number of iterations did not affect the overall inference of $K$. We 
observed that often during the last iterations individuals would alternate between 
reference groups. Increasing the log likelihood 
difference threshold for reallocation to 0.18 (requiring at least a 1.5 times 
higher likelihood to reallocate the individual) minimized this, but did not affect the 
inference of $K$ (lower bound estimate of $K = 2$).

\section*{Conclusions}
In the past decade, the use of genetic markers to identify population structure and to attempt inference of
the number of genetic clusters ($K$) in a collection has dramatically increased in the fields of ecology, evolution, epidemiology, and conservation
genetics. By virtute of this expanding desire to identify $K$ there has been a push to 
develop fast and easy-to-use methods that can accurately infer $K$. We show that {\sc flock}, a program billed as providing
``more accurate allocations and more reliable estimates of $K$ [while running]
much faster than {\sc structure} \citep[][p.~734]{Duc&Tur2012},'' may not, in fact, represent such 
a radical departure from {\sc structure} itself.  In particular, we have shown that the algorithm implemented 
in {\sc flock} is a simulated annealing algorithm for the Bayesian 
MAP estimate from a marginalized and restricted form of the {\sc structure} model.  
Given such similarities between the two methods it is not surprising that
we observed relatively
congruent results between the ``best run'' from {\sc flock} and {\sc structure} using 
a real data set from \textit{O. mykiss}.

Though we observed very similar results
between the two programs at low values of $K$, {\sc flock} had a tendency to provide inconsistent solutions
among its ``best runs'' at higher values, while still finding solutions that were similar to {\sc structure}'s.
This is likely a consequence of the rapidly increasing size of the search space over possible 
partitions as $K$ and $n$, the number of individuals, increase. 
The likelihood surface defined on partitions of this space is likely to be
be characterized by multiple peaks and troughs.
The peak or plateau that  {\sc flock}'s deterministic algorithm finds is entirely dependent on the initial allocation of 
individuals, as it has no way to visit areas of the surface with smaller likelihood values.
If the number of initial allocations does not increase
as $K$ is increased, the starting allocations become diffusely scattered in the space of all
possible partitions making it unlikely that they will converge. This coupled with the fact that there is 
ambiguity in what defines a "best run" when all plateaus are of length zero complicates the interpretation
of results from {\sc flock} if no non-zero plateaus are encountered.

One of the major claimed advantages of {\sc flock} over other methods for genetic clustering (like {\sc structure})
is that, ``The very short processing times for each run of {\sc flock} allows for
comparing partitions from many runs of each $K$'' \citep[][p.~735]{Duc&Tur2012}.
While the authors observed faster processing times (relative to other methods) in their 
comparisons, we did not observe the same advantage in our analysis. This appears to result from two issues. First,
our example data set was an order of magnitude larger than the largest data set analyzed in \citet{Duc&Tur2012}.  Second, 
in the comparison by \citet{Duc&Tur2012}, 
{\sc flock} was set to perform 50 runs and 20 iterations for each $K$ from 1 to four more than 
the true value of $K$, and computing times for that analysis were compared to times for ten iterations of {\sc structure} using 
a 50,000 sweep burn-in followed 
by a 200,000 sweep sample from the posterior of $K$ from 1 to three more than the true 
value of $K$. 
To us it seemed reasonable to compare, for each $K$, one iteration of {\sc structure} (50,000 burn in and 200,000 sample collection sweeps)
to one iteration of {\sc flock} (50 different starting allocations that are reallocated 20 times) 
as both are used to provide a single estimate of the partition of individuals into clusters. One could
argue that it would be more appropriate to compare the running times of the programs according to the number
of opportunities that each individual's cluster membership has for random initialization and reallocation.  Under 
such a perspective, {\sc flock}'s 50 different starting allocations and 20 reallocations should be compared to
50 separate runs of structure, each of length 20 sweeps, which would surely give {\sc structure}
an apparent advantage in speed.  Regardless, while it may hold that {\sc flock}
is able to process small datasets quickly enough to compete with {\sc structure},
the rapid increase in our ability to economically
genotype a large number of individuals for a great many loci, may render this advantage less relevant to many 
study designs in the future.

For those that do decide to use {\sc flock} as an alternative to {\sc structure}, the adhoc rules 
to estimate $K$ may be overly stringent. While there has been lengthy 
discourse over the choice of estimators for $K$ in {\sc structure} \citep{Pritchardetal2000,Evannoetal2005,Wap&Gag2006,Gaoetal2011} much less effort has been put forth examining how the number of individuals
and the number of starting allocations can affect not only the convergence of different runs of {\sc flock}, but also its 
inference of $K$. The authors developed estimation rules based on simulated data 
sets with a modest number of individuals and different migration regimes and rates, number of loci, and $K$. 
By relying on identical allocations (mean LLOD values) as a means of supporting 
a particular $K$,  similar---but not identical---allocations
will provide no support to a given $K$. If using {\sc flock}, we advise running
multiple small batches and using the programs {\sc clumpp} \citep{Jak&Ros2007} and
{\sc distruct} \citep{Rosenberg2004} to visually inspect the ``best run'' results.

For many types of genetic analyses, molecular ecologists can choose from 
amongst a wide range of published methods.  Choosing one of these methods to use can
be difficult, especially when the methods have been touted as very different.
We have known colleagues whose solution to this problem is simply to run their data
through every program available.  This seems an ineffective use of time and resources 
that could be counteracted by a more transparent accounting of the similarities between
methods. As one small step toward that,   
we have provided a different prespective on the algorithm
implemented in {\sc flock}, showing it to be very similar to {\sc structure}
using a particular set of options.  We hope this might increase understanding of {\sc flock}'s applicability to different datasets while decreasing end user confusion and misuse.

>>>>

\subsection*{{\sc flock} as a special case of {\sc structure}}
The goal of {\sc structure}'s MCMC algorithm is to sample from the space of 
$Z$'s in proportion to their posterior probability.  If, all that was desired
was a point estimate of the $Z_i$'s that maximized the posterior probability,
then one standard approach to finding that maximum {\em a posteriori}, or MAP, 
estimate, would be to use simulated annealing \citep{Kirkpatricketal1983}.
To do simulated annealing in the {\sc structure} model with no admixture and
no correlated allele frequency prior, after integrating out the unknown
allele frequencies, the updates for each $Z_i$ would be made from the 
distribution
\[
\biggl[P(Z_i=k~|~Y_i, ~Z_{(-i)},~Y_{(-i)\ell})\biggr]^\beta
\]
which is the same full conditional distribution as the marginalized {\sc structure}
model (equation~\ref{eq:fc}), raised
to the power $\beta$.  In typical applications of simulated annealing,
$\beta$ is set to a starting value less than 1, and, as the algorithm proceeds,
$\beta$ is increased according to a ``cooling schedule''\citep{Hajek1988}, until
the algorithm finds a local maximum in the posterior probability surface.  
When $\beta$ is small, the process can easily traverse ``valleys''
in the posterior probability surface and has a better chance of searching
more of the space for the maximum value.  As $\beta$ gets larger, it becomes
more probable that the process will
move uphill on the probability surface, tending toward the local
maximum near where it currently is.

@@

For each set of runs, {\sc flock} reports the 
normalized likelihood values (the exponentiated log-likelihood value for an individual's 
membership to each of $K$ clusters, scaled to sum to one) for a run belonging to the longest plateau - refered to as the 'best run' by \citep{Duc&Tur2012} (see below).
These values can be compared to the $q_i$ values
returned by {\sc structure}, which, in the no-admixture case, are posterior probabilities of cluster membership.
The program {\sc clumpp} \citep{Jak&Ros2007} was used to relabel clusters among different runs,
while the program {\sc district} \citep{Rosenberg2004} was used to plot individual \textit{$q_i$} values.

\citet{Duc&Tur2012} promote a method for using {\sc flock} to estimate, $K$, the number of
subpopulations. The user must specify both the number of runs (different initial allocations of individuals 
to each cluster) as well as the number of iterations. (In each iteration every individual has the opportunity to be reallocated 
to a different cluster.) At the conclusion of each run, the log-likelihood value for membership to all $K$ clusters is computed
and the log likelihood difference (LLOD), the difference between the cluster with the highest likelihood 
and second highest likelihood, is calculated for each individual. {\sc flock} identifies runs that 
have converged to the same solution by detecting that they have the same mean LLOD score, and, in 
{\sc flock} parlance, a series of runs that have converged to the same solution are termed a ``plateau.'' 
The plateau record is a list of how many runs converged to the same allocation
of individuals. 
\citet{Duc&Tur2012} developed two stopping rules that they use to estimate $K$. In the absence of 
strong prior beliefs about the value of $K$, \citet{Duc&Tur2012} advocate running  {\sc flock} until either a single plateau of
$\geq 6$ runs is reached or four consecutive $K$ values yield no plateaus. If the first stopping condition is satisfied,
the $K$ for which there is a single plateau of $\geq 6$ runs is the point estimate for $K$. Alternatively, 
if multiple plateaus are observed with at least one plateau of $\geq 6$ runs, the largest value of $K$ that satisfies 
this conditon should be taken as the lower bound estimate of $K$. If no plateaus of $\geq 6$ runs are observed, $K$
is declared {\em undecided}.

@@

\subsection*{Results} 
Run times for both programs were relatively short, albeit not trivial. All runs of {\sc flock} 
completed in 1133 minutes. Each of the six groups of runs averaged 188.8 minutes.
All runs of {\sc structure} were completed in 
486 minutes and averaged 80.8 minutes to complete each of the six groups of runs.
The processing time of each algorithm increased as the number or 
putative clusters increases. Increasing
$K$ from of 2 to 6 increased the run time for {\sc flock} by a factor of 2.4, while the run time 
for {\sc structure} increased by a factor of 1.6.

{\sc flock} and {\sc structure} provide similar results in 
terms of how individuals are allocated among runs at low values of $K$, 
while at larger values, {\sc flock} often found, in addition to the solution identified by {\sc structure}, 
multiple other solutions. For all six groups 
of runs with $K = 2$ to 4, {\sc flock} obtained a nearly identical clustering of individuals 
(See supplementary materials) none of which differed substantially from {\sc structure}'s results
(Fig.~\ref{fig:qplots}). Of the ``best runs'' for the six groups for $K$ = 2, two  
runs yielded identical clustering (same mean LLOD score) while all other runs 
differed by at most the allocation of 13 individuals. As 
$K$ increased so did the number of solutions
that {\sc flock} found (Fig 1). For all groups of runs for $K$ $\geq 2$, no plateaus were observed---all intial 
allocations resulted in solutions that differed. 
{\sc flock} still produced estimates of the LLOD by individual for the 
'best run', however, it is unclear in these cases what constitutes the 'best run' as all plateau lengths are 
equal to one. Many of these solutions are, however, surprisingly similar to
the partitions identified by {\sc structure}. Both programs idenfity clusters 
which can be attributed to the geographical boundaries identified by \citet{Garzaetal_norcal},
although less reliably so in the case of {\sc flock}.

@@

To investigate the ability of the arrangement of individuals within the reference 
populations to move around the partition space we varied the number of iterations that 
each run goes through to find the ``optimal'' allocation of individuals. Unfortuantely 
only the reallocation matrix for the ``best run'' is given as output. Increasing the 
number of iterations did not affect the overall inference of $K$. We 
observed that often during the last iterations individuals would alternate between 
reference groups. Increasing the log likelihood 
difference threshold for reallocation to 0.18 (requiring at least a 1.5 times 
higher likelihood to reallocate the individual) minimized this, but did not affect the 
inference of $K$ (lower bound estimate of $K = 2$).

\section*{Conclusions}
In the past decade, the use of genetic markers to identify population structure and to attempt inference of
the number of genetic clusters ($K$) in a collection has dramatically increased in the fields of ecology, evolution, epidemiology, and conservation
genetics. By virtue of this expanding desire to identify $K$ there has been a push to 
develop fast and easy-to-use methods that can accurately infer $K$. We show that {\sc flock}, a program billed as providing
``more accurate allocations and more reliable estimates of $K$ [while running]
much faster than {\sc structure} \citep[][p.~734]{Duc&Tur2012},'' may not, in fact, represent such 
a radical departure from {\sc structure} itself.  In particular, we have shown that the algorithm implemented 
in {\sc flock} is a simulated annealing algorithm for the Bayesian 
MAP estimate from a marginalized and restricted form of the {\sc structure} model.  
Given such similarities between the two methods it is not surprising that
we observed relatively
congruent results between the ``best run'' from {\sc flock} and {\sc structure} using 
a real data set from \textit{O. mykiss}.

Though we observed very similar results
between the two programs at low values of $K$, {\sc flock} had a tendency to provide inconsistent solutions
among its ``best runs'' at higher values of $K$, while still finding solutions that were similar to {\sc structure}'s.
This is likely a consequence of the rapidly increasing size of the search space over possible 
partitions as $K$ and $n$, the number of individuals, increase. 
The likelihood surface defined on partitions of this space is likely to be
be characterized by multiple peaks and troughs.
The peak that  {\sc flock}'s deterministic algorithm finds is entirely dependent on the initial allocation of 
individuals, as it has no way to visit areas of the surface with smaller likelihood values.
If the number of initial allocations does not increase
as $K$ is increased, the starting allocations become diffusely scattered in the space of all
possible partitions making it unlikely that many (if any at all) will converge to the same local mode. This coupled with the fact that there is 
ambiguity in what defines a ``best run'' when all plateaus are of length zero complicates the interpretation
of results from {\sc flock} if no non-zero plateaus are encountered.

One of the major claimed advantages of {\sc flock} over other methods for genetic clustering (like {\sc structure})
is that, ``The very short processing times for each run of {\sc flock} allows for
comparing partitions from many runs of each $K$'' \citep[][p.~735]{Duc&Tur2012}.
While \citeauthor{Duc&Tur2012} observed faster processing times (relative to other methods) in their 
comparisons, we did not observe the same advantage in our analysis. This appears to result from two issues. First,
our example data set was an order of magnitude larger than the largest data set analyzed in \citet{Duc&Tur2012}.  Second, 
in the comparison by \citet{Duc&Tur2012}, 
{\sc flock} was set to perform 50 runs and 20 iterations for each $K$ from 1 to four more than 
the true value of $K$, and computing times for that analysis were compared to times for ten iterations of {\sc structure} using 
a 50,000 sweep burn-in followed 
by a 200,000 sweep sample from the posterior for values of $K$ from 1 to three more than the true 
value of $K$. 
To us it seemed reasonable to compare, for each $K$, one iteration of {\sc structure} (50,000 burn in and 200,000 sample collection sweeps)
to one iteration of {\sc flock} (50 different starting allocations that are reallocated 20 times) 
as both are used to provide a single estimate of the partition of individuals into clusters. One could
argue that it would be more appropriate to compare the running times of the programs according to the number
of opportunities that each individual's cluster membership has for random initialization and reallocation.  Under 
such a perspective, {\sc flock}'s 50 different starting allocations and 20 reallocations should be compared to
50 separate runs of structure, each of length 20 sweeps, which would surely give {\sc structure}
an apparent advantage in speed.  Regardless, we did not find a speed advantage from using {\sc flock}.
If anything, {\sc structure} was somewhat faster.

For those that do decide to use {\sc flock} as an alternative to {\sc structure}, the {\em ad hoc} rules 
to estimate $K$ may be overly stringent. While there has been lengthy 
discourse over the choice of estimators for $K$ in {\sc structure} \citep{Pritchardetal2000,Evannoetal2005,Wap&Gag2006,Gaoetal2011} much less effort has been put forth examining how the number of individuals
and the number of starting allocations can affect not only the convergence of different runs of {\sc flock}, but also its 
inference of $K$. \citeauthor{Duc&Tur2012} developed estimation rules based on simulated data 
sets with a modest number of individuals and different migration regimes and rates, number of loci, and $K$. 
By relying on identical allocations (mean LLOD values) as a means of supporting 
a particular $K$,  similar---but not identical---allocations
will provide no support to a given $K$. If using {\sc flock}, we advise running
multiple small batches and using the programs {\sc clumpp} \citep{Jak&Ros2007} and
{\sc distruct} \citep{Rosenberg2004} to visually inspect the ``best run'' results.

For many types of genetic analyses, molecular ecologists can choose from 
amongst a wide range of published methods.  Choosing one of these methods to use can
be difficult, especially when the methods have been touted as very different.
We have known colleagues whose solution to this problem is simply to run their data
through every program available.  This seems an ineffective use of time and resources 
that could be counteracted by a more transparent accounting of the similarities between
methods. As one small step toward that,   
we have provided a different perspective on the algorithm
implemented in {\sc flock}, showing it to be very similar to {\sc structure}
using a particular set of options.  We hope this might increase understanding of {\sc flock}'s applicability to different datasets while decreasing end user confusion and misuse.
